Practical Optional Types for Clojure

### 2015-02-20 formal-model.tex
HMap prose

\begin{figure}
  \footnotesize
\begin{mathpar}
  \Multisubtyping{}
\end{mathpar}
\caption{Multimethod subtyping rules}
\end{figure}

@@

The typing rules (figure~\ref{figure:mmtyping}) utilse the {\MultiFntype{\s{}}{\t{}}}
type, where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies.

@@

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap subtyping rules}
\end{figure}

@@

The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
HMap can assume all entries not in {\mandatory{}} as absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only allow the empty map literal and
restrict lookup and extension to keyword keys.

>>>>

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \caption{Multimethod subtyping rules}
  \label{figure:mmsubtype}
\end{figure}

@@

The typing rules (figure~\ref{figure:mmtyping}) use the multimethod type {\MultiFntype{\s{}}{\t{}}}, 
where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies. T-App handles beta reduction because multimethods are subtypes of functions
(figure~\ref{figure:mmsubtype}).

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\BAssoc}

    {\BGet}

    {\BGetMissing}
  \end{mathpar}
  \caption{HMap Semantics}
  \label{figure:hmapsem}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap subtyping rules}
  \label{figure:hmapsubtype}
\end{figure}

@@

The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we write \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}} as a fully specified
HMap can assume all entries not in {\mandatory{}} as absent. We disallow keys to be both present and absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only allow the empty map literal and
restrict lookup and extension to keyword keys. The runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}
is not part of the language syntax.

Figure~\ref{figure:hmapsem} shows the semantics for HMaps, where B-Get returns the associated
entry if it is present and B-GetMissing returns \nil{} otherwise. B-Assoc returns a new extended
map.

Typing rules are given in figure~\ref{figure:hmaptyping}. T-GetHMap handles the simple
case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
a \Nil{} return type.
T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
T-Assoc extends the known keyword entries, overwriting previous entries with the same key.

S-HMap (figure~\ref{figure:hmapsubtype}) says that HMap subtypes must have all present entries to also be
subtypes and also cover the absent keys.
Complete maps are subtypes of partial maps by S-HMapP.
